---
title: "Kidney_CB2_Project"
author: "Daire Scully"
date: "2026-02-11"
output: html_document
---


```{r}
# Packages needed in this section
req_pkgs <- c(
  "dplyr",      # data wrangling
  "ggplot2",    # plotting
  "tidyr",      # tidying
  "readr",      # read/write csv
  "tibble",     # tibbles/printing
  "gridExtra",  # simple plot grids
  "emmeans",    # adjusted means / contrasts
  "effsize",     # effect sizes (Cohen's d, Cliff's delta)
 "DataExplorer" ,#for missingness
  "glmnet",
 "lmtest" ,# for checking the 
 "pROC"
 )

# Install any that are missing
to_install <- setdiff(req_pkgs, rownames(installed.packages()))
if (length(to_install) > 0) {
  install.packages(to_install, dependencies = TRUE)
}
# Load all (silently)
invisible(lapply(req_pkgs, require, character.only = TRUE))

```


```{r}
#set working directory and acknowledge dataset
getwd()
setwd("C:/Users/pbsstudent.L-1WTXKR3/Desktop/R Projects/CB2/Project 1 - Kidney/CB2_first")
file.exists("C:/Users/pbsstudent.L-1WTXKR3/Desktop/R Projects/CB2/Project 1 - Kidney/CB2_first/dataset02_kidney_FULL.csv")
```

```{r}
#setting kidney_full as the dataset name for R
kidney_full <- read.csv("C:/Users/pbsstudent.L-1WTXKR3/Desktop/R Projects/CB2/Project 1 - Kidney/CB2_first/dataset02_kidney_FULL.csv")
head(kidney_full) #as less than 15 columns
str(kidney_full[, 1:10])
str(kidney_full)
```

```{r}
#defining low and high adherance thersholds
kidney_full <- kidney_full %>%
  dplyr::mutate(
    adherence_level = dplyr::case_when(
      protocol_adherence >= 0.6 ~ "High adherence",
      protocol_adherence <  0.6 ~ "Low adherence",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::mutate(
    adherence_level = factor(
      adherence_level,
      levels = c("Low adherence", "High adherence")
    )
  )
head(kidney_full)
```

```{r}
kidney_full <- kidney_full %>%
  dplyr::mutate(
    creatinine_change = dplyr::case_when(
      delta_creatinine_48h >= 0 ~ "Creat increased",
      delta_creatinine_48h <  0 ~ "Creat reduced",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::mutate(
    creatinine_change = factor(
      creatinine_change,
      levels = c("Creat reduced", "Creat increased")
    )
  )
head(kidney_full)
```


#second part
```{r}
#Checking for missing variables and visualising:

type_map <- tibble::tibble(
  variable    = names(kidney_full),
  class       = sapply(kidney_full, \(x) paste(class(x), collapse = "/")),
  n_missing   = sapply(kidney_full, \(x) sum(is.na(x))),
  pct_missing = round(100 * sapply(kidney_full, \(x) mean(is.na(x))), 2),
  n_unique    = sapply(kidney_full, \(x) dplyr::n_distinct(x)),
  example     = sapply(kidney_full, \(x) paste(utils::head(unique(x), 3), collapse = ", "))
)
library(dplyr)
type_map %>% arrange(desc(class))
library(ggplot2)
ggplot(type_map %>% arrange(desc(pct_missing)), aes(x = variable, y = pct_missing)) +
  geom_col() + scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, .05)))
```

Exploratory analysis and visualisation:

Creatinine is a waste product which is produced by normal muscle breakdown and glomerular filtration.Levels are stable in homeostatic conditions, it accumulates in the blood during decline of kidney function. Rises in creatinine can be an early indicator of kidney issues allowing earlier intervention by HCPs. It's use in this dataset allows us to see both long term and short term condition changes due to the frequency at which it is monitored.

Normal creatinine levels are generally 0.7 to 1.3 mg/dL for men and 0.6 to 1.1 mg/dL for women, reflecting healthy kidney function. Levels above 1.4 mg/dL (men) or 1.2 mg/dL (women) often indicate kidney dysfunction. A result of 1.6–2.0 mg/dL is considered moderately elevated, while levels 5.0 mg/dL may indicate severe damage.

```{r}
#plots showing the counts of each variable in histogram and bar format
p1 <- ggplot(kidney_full, aes(delta_creatinine_48h)) + geom_histogram(bins = 30) +
  labs(x = "Percentage Change", y = "Count")
p2 <- ggplot(kidney_full, aes(x = ifelse(bun < 20, "< 20", "≥ 20"))) + geom_bar() +
  labs(x = "High BUN (≥20%)", y = "Count")
p3 <- ggplot(kidney_full, aes(baseline_creat)) + geom_histogram(bins = 30) +
  labs(x = "Baseline Creatinine", y = "Count")
p4 <- ggplot(kidney_full, aes(age)) + geom_histogram(bins = 30) +
  labs(x = "Age (years)", y = "Count")
p5 <- ggplot(kidney_full, aes(protocol_adherence)) + geom_histogram(bins = 30) +
  labs(x = "Protocol Adherence", y = "Count")
p6 <- ggplot(kidney_full, aes(sofa)) + geom_histogram(bins = 30) +
  labs(x = "SOFA Score", y = "Count")
p7 <- ggplot(kidney_full, aes(site)) + geom_bar() +
  labs(x = "ICU Site", y = "Count")
p8 <- ggplot(kidney_full, aes(x = ifelse(protocol_adherence < 0.7, "< 70%", "≥ 70%"))) + geom_bar() +
  labs(x = "High Adherence (≥70%)", y = "Count")
gridExtra::grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, ncol = 2)
```

```{r}
#plots comparing variables to eachother
ggplot(kidney_full, aes(baseline_creat, delta_creatinine_48h, color = sofa)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  labs(x = "Baseline Creatinine", y = "Delta Creatinine", color = "SOFA score")

ggplot(kidney_full, aes(delta_creatinine_48h, protocol_adherence, color = age)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  labs(x = "Creatinine after 48h", y = "CRP", color = "age")

ggplot(kidney_full, aes(baseline_creat, delta_creatinine_48h, color = protocol_adherence)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  labs(x = "Baseline Creatinine", y = "Delta Creatinine", color = "Protocol Adherence")

ggplot(kidney_full, aes(baseline_creat, protocol_adherence, color = delta_creatinine_48h < 0 )) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  labs(x = "Baseline Creatinine", y = "Protocol Adherence", color = "Creatinine after 48h less than 0")
```


```{r}
#examining relationship between sofa >7 on variable expression

ggplot(kidney_full, aes(sofa > 7, delta_creatinine_48h)) +
  geom_boxplot() +
  labs(x = "SOFA Score >7", y = "Delta Creatinine (after 48h)")

ggplot(kidney_full, aes(sofa > 7, delta_creatinine_48h)) +
  geom_boxplot(outlier.shape = NA, width = 0.6) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white") +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.15) +
  labs(x = "SOFA Score >7", y = "Creatinine after 48h")

ggplot(kidney_full, aes(protocol_adherence > 0.8, delta_creatinine_48h)) +
  geom_boxplot() +
  labs(x = "Protocol Adherence > 80%", y = "Delta Creatinine (after 48h)")
```

Summarising the Dataset:

```{r}
#statistical analysis of each sofa scoring and adherance group (n, mean, sd, median and q1 & 3)
kidney_full %>%
  group_by(sofa) %>%
  summarise(
    n      = n(),
    mean   = mean(baseline_creat),
    sd     = sd(baseline_creat),
    median = median(baseline_creat),
    q1     = quantile(baseline_creat, 0.25),
    q3     = quantile(baseline_creat, 0.75)
  )

kidney_full %>%
  group_by(protocol_adherence) %>%
  summarise(
    n      = n(),
    mean   = mean(delta_creatinine_48h),
    sd     = sd(delta_creatinine_48h),
    median = median(delta_creatinine_48h),
    q1     = quantile(delta_creatinine_48h, 0.25),
    q3     = quantile(delta_creatinine_48h, 0.75)
  )

kidney_full %>%
  group_by(creatinine_change) %>%
  summarise(
    n      = n(),
    mean   = mean(baseline_creat),
    sd     = sd(baseline_creat),
    median = median(baseline_creat),
    q1     = quantile(baseline_creat, 0.25),
    q3     = quantile(baseline_creat, 0.75)
  )

```

Determining Correlation:

CRP:C-Reactive Protein is produced by the liver in response to inflammation. It is acute phase reactant (inflammation, infection or tissue damage) and a highly sensitive marker. It is used to monitor the presence and severity of inflammatory processes. 

In the particular cohort we are dealing with here, CRP is particularly valuable.Inflammatory responses in ICU can be common indicators of sepsis, trauma, surgery or organ failure and rising or persistently high CRP levels can be an indicator of lack of treatment response. CRP is measured through routine blood tests making it also accessible and practical for serial monitoring at bedside settings. In our analysis CRP acts as an important clinical marker for understanding how inflammatory burden interacts with kidney function and other factors in the patient group.

Values: with a normal level generally below 3-10 mg/L (or 1mg/dL). Levels between 10-100 mg/L indicate moderate inflammation (infection, chronic disease), while levels >100 mg/L suggest severe inflammation or acute bacterial infection.

```{r}
lab_markers <- c("bun", "crp", "albumin", "lactate", "map_min", "sofa")
cors <- sapply(lab_markers, function(g) {
  suppressWarnings(cor(kidney_full[[g]], kidney_full$baseline_creat, use = "pairwise.complete.obs"))
})
top_marker <- names(sort(abs(cors), decreasing = TRUE))
top_mark <- tibble::tibble(
  Marker = top_marker,
  cor  = unname(cors[top_marker])
) %>%
  dplyr::mutate(Marker = reorder(Marker, abs(cor)))

ggplot(top_mark, aes(x = Marker, y = cor, fill = cor > 0)) +
  geom_col() +
  coord_flip() +
  guides(fill = "none") +
  labs(x = "Laboratory Markers", y = "Correlation with Baseline Creatinine",
       title = "Top Lab Markers by absolute correlation")
```

```{r}
Proto <- c("protocol_adherence")
correl <- sapply(Proto, function(g) {
  suppressWarnings(cor(kidney_full[[g]], kidney_full$delta_creatinine_48h, use = "pairwise.complete.obs"))
})

top_mark <- tibble::tibble(
  Mark = correl,
  corr  = unname(cors[correl])
) %>%
  dplyr::mutate(Mark = reorder(Mark, abs(corr)))

ggplot(top_mark, aes(x = Mark, y = corr, fill = corr > 0)) +
  geom_col() +
  coord_flip() +
  guides(fill = "none") +
  labs(x = "Protocol Adherence", y = "Correlation with Delta Creatinine",
       title = "Protocol Adherence by absolute correlation")
```

Linear Regression:

```{r}
ggplot(kidney_full, aes(x = baseline_creat, y = delta_creatinine_48h, colour = protocol_adherence)) +
geom_point(alpha = 0.7) +
geom_smooth(method = "lm", se = FALSE, color = "black") +
labs( x = "Baseline Creatinine", y = "Delta Creatinine (48h)", color = "protocol adherence") +
theme_minimal()

ggplot(kidney_full, aes(x = baseline_creat, y = delta_creatinine_48h, colour = age)) +
geom_point(alpha = 0.7) +
geom_smooth(method = "lm", se = FALSE, color = "black") +
labs( x = "Baseline Creatinine", y = "Delta Creatinine (48h)", color = "Age") +
theme_minimal()

ggplot(kidney_full, aes(x = crp, y = delta_creatinine_48h, colour = protocol_adherence)) +
geom_point(alpha = 0.7) +
geom_smooth(method = "lm", se = FALSE, color = "red") +
labs( x = "CRP", y = "Delta Creatinine (48h)", color = "protocol adherence") +
theme_minimal()

ggplot(kidney_full, aes(x = baseline_creat, y = protocol_adherence, colour = delta_creatinine_48h < 0 )) +
geom_point(alpha = 0.7) +
geom_smooth(method = "lm", se = FALSE, color = "green") +
labs( x = "Baseline Creatinine", y = "Protocol Adherence", color = "Creatinine change after 48h") +
theme_minimal()
```

Calculating Error:

```{r}
# ---- 0) Metrics (MAE and RMSE) ----

mae  <- function(y, yhat) mean(abs(y - yhat))
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))

# ---- evaluation function (MAE and RMSE) ----
eval_perf <- function(y_true, y_pred) {
  tibble(
    MAE  = mae(y_true, y_pred),
    RMSE = rmse(y_true, y_pred)
  )
}
# alias to avoid mistyping
eval_perform <- eval_perf
```

Splitting Data into Train and Test

```{r}
set.seed(42)
# ---- 1) Train/Test split (70/30) ----
n  <- nrow(kidney_full)
ix <- sample.int(n, size = floor(0.7 * n))
train <- kidney_full[ix, , drop = FALSE]
test  <- kidney_full[-ix, , drop = FALSE]

# ---- 2) Remove columns that must NOT enter the model ----
# (ID and the binary endpoint; keeps the supervised task as a pure regression)

drop_cols <- intersect(names(train), c("patient_id", "protocol_adherence", "site"))
train_nopii <- dplyr::select(train, -dplyr::all_of(drop_cols))
test_nopii  <- dplyr::select(test,  -dplyr::all_of(drop_cols))

# Common formula for OLS (and to derive terms/dummies for glmnet)
f_ols <- delta_creatinine_48h ~ .

# ---- 3) Capture TRAIN terms to ensure identical dummies in TEST ----
# (fit a temporary OLS only to extract terms & factor levels)

ols_tmp   <- lm(f_ols, data = train_nopii)
ols_terms <- terms(ols_tmp)

# Consistent design matrices for glmnet (no intercept column)
X_train <- model.matrix(ols_terms, data = train_nopii)[, -1, drop = FALSE]
X_test  <- model.matrix(ols_terms, data = test_nopii)[,  -1, drop = FALSE]
y_train <- train_nopii$delta_creatinine_48h
y_test  <- test_nopii$delta_creatinine_48h

# ---- 4) OLS (uses data.frame; glmnet uses X/y) ----
mod_ols        <- lm(f_ols, data = train_nopii)
pred_ols_train <- predict(mod_ols, newdata = train_nopii)
pred_ols_test  <- predict(mod_ols, newdata = test_nopii)

  #| label: ols-assumptions
  #| message: false
  #| warning: false
  suppressPackageStartupMessages({
    library(ggplot2)
    library(dplyr)
    library(lmtest)    # bptest(), dwtest()
    library(sandwich)  # robust (HC) variance estimators
    library(car)       # vif(), crPlots(), outlierTest()
  })
  
  # --- 1) Quick base-R diagnostic panel (residuals, QQ, Scale-Location, Residuals vs Leverage)
op <- par(mfrow = c(2, 2))
plot(mod_ols)
```

OLS:

```{r}
# --- Evaluate OLS with MAE and RMSE (using eval_perf) ---
# y vectors (true values)
y_train <- train_nopii$delta_creatinine_48h
y_test  <- test_nopii$delta_creatinine_48h

perf_ols_train <- eval_perf(y_train, pred_ols_train) |>
  dplyr::mutate(Model = "OLS", Dataset = "Train")

perf_ols_test <- eval_perf(y_test, pred_ols_test) |>
  dplyr::mutate(Model = "OLS", Dataset = "Test")

# Compact table
dplyr::bind_rows(perf_ols_train, perf_ols_test)
```

LASSO Regression:

```{r}
# ---- 5) LASSO (alpha = 1) with CV to choose lambda ----
# Note: glmnet standardizes features by default (standardize = TRUE).
library(glmnet)
cv_lasso  <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 10)
mod_lasso <- glmnet(X_train, y_train, alpha = 1, lambda = cv_lasso$lambda.min)
pred_lasso_train <- as.numeric(predict(mod_lasso, newx = X_train))
pred_lasso_test  <- as.numeric(predict(mod_lasso, newx = X_test))

# Nonzero coefficients selected by LASSO
coef_lasso <- coef(mod_lasso)
nz <- which(coef_lasso != 0)
as.matrix(coef_lasso[nz, , drop = FALSE])

# Compare errors with your eval_perf()
perf_lasso_train <- eval_perf(y_train, pred_lasso_train) |> dplyr::mutate(Model="LASSO", Dataset="Train")
perf_lasso_test  <- eval_perf(y_test,  pred_lasso_test)  |> dplyr::mutate(Model="LASSO", Dataset="Test")
dplyr::bind_rows(perf_lasso_train, perf_lasso_test)
```

Ridge Regression:

As part of our analysis we used a ridge regression. This is a regularized linear modelling technique that's particularly well suited for data sets made up of interrelated variables such as those used to assess kidney function. Unlike standard linear regression a penalty is applied which shrinks all coefficient estimates toward 0 but doesn't eliminate them entirely. In this example it was able to do so for creatinine, BUN, lactate and other lab markers.

By keeping all of these variables in the model while shrinking their influence we can reduce overfitting so that our findings will be suitable for future patients and stabilizes coefficient estimates even thoguh there are correlations between the our predictors. This gives us a more reliable image of which factors are most strongly associated without having to remove potentially informative variables.

```{r}
# ---- 6) Ridge (alpha = 0) with CV ----
cv_ridge  <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 10)
mod_ridge <- glmnet(X_train, y_train, alpha = 0, lambda = cv_ridge$lambda.min)
pred_ridge_train <- as.numeric(predict(mod_ridge, newx = X_train))
pred_ridge_test  <- as.numeric(predict(mod_ridge, newx = X_test))

# ---- Performance (MAE, RMSE) using eval_perform / eval_perf ----
perf_ridge_train <- eval_perform(y_train, pred_ridge_train) |>
  dplyr::mutate(Model = "Ridge", Dataset = "Train")

perf_ridge_test <- eval_perform(y_test, pred_ridge_test) |>
  dplyr::mutate(Model = "Ridge", Dataset = "Test")

dplyr::bind_rows(perf_ridge_train, perf_ridge_test)

# ---- Quick model “summary” for discussion ----
# Lambdas chosen by CV
cv_ridge$lambda.min

cv_ridge$lambda.1se

# Coefficients at lambda.min
ridge_coefs <- as.matrix(coef(mod_ridge))
ridge_coef_tbl <- tibble::tibble(
  term     = rownames(ridge_coefs),
  estimate = as.numeric(ridge_coefs[, 1])
)

# How many non-zero coefficients (excluding intercept)
nnz <- ridge_coef_tbl |>
  dplyr::filter(term != "(Intercept)", estimate != 0) |>
  nrow()

nnz

# Top coefficients by absolute magnitude (exclude intercept)
ridge_top <- ridge_coef_tbl |>
  dplyr::filter(term != "(Intercept)") |>
  dplyr::mutate(abs_est = abs(estimate)) |>
  dplyr::arrange(dplyr::desc(abs_est)) |>
  dplyr::slice(1:15)
## seeing the top 15 features regarding estimate
ridge_top
```

Elastic Net:

```{r}
# ---- 7) Elastic Net (tune alpha and lambda via CV) ----
suppressPackageStartupMessages({
  library(glmnet)
  library(dplyr)
  library(tibble)
  # assumes eval_perform() already defined and X_train, X_test, y_train, y_test exist
})

set.seed(42)

# Grid of alpha values (0=ridge, 1=lasso)
alpha_grid <- seq(0.05, 0.95, by = 0.05)

# For each alpha, run cv.glmnet to pick lambda; record CV error at lambda.min
cv_list <- lapply(alpha_grid, function(a) {
  cv.glmnet(X_train, y_train, alpha = a, nfolds = 10)
})

cv_errors <- sapply(cv_list, function(cv) min(cv$cvm))  # lower is better
best_idx  <- which.min(cv_errors)
best_alpha <- alpha_grid[best_idx]
best_cv    <- cv_list[[best_idx]]
best_lambda_min <- best_cv$lambda.min
best_lambda_1se <- best_cv$lambda.1se

best_alpha

best_lambda_min

best_lambda_1se

# Fit final Elastic Net at best alpha/lambda
mod_enet <- glmnet(X_train, y_train, alpha = best_alpha, lambda = best_lambda_min)

# Predictions
pred_enet_train <- as.numeric(predict(mod_enet, newx = X_train))
pred_enet_test  <- as.numeric(predict(mod_enet, newx = X_test))

# Performance
perf_enet_train <- eval_perform(y_train, pred_enet_train) |>
  mutate(Model = "Elastic Net", Dataset = "Train", alpha = best_alpha, lambda = best_lambda_min)

perf_enet_test <- eval_perform(y_test, pred_enet_test) |>
  mutate(Model = "Elastic Net", Dataset = "Test", alpha = best_alpha, lambda = best_lambda_min)

bind_rows(perf_enet_train, perf_enet_test)

# Coefficients summary
enet_coefs <- as.matrix(coef(mod_enet))
enet_coef_tbl <- tibble(
  term     = rownames(enet_coefs),
  estimate = as.numeric(enet_coefs[, 1])
)

# Count non-zero (excluding intercept)
enet_nnz <- enet_coef_tbl |>
  filter(term != "(Intercept)", estimate != 0) |>
  nrow()
enet_nnz

# Top coefficients by absolute value (exclude intercept)
enet_top <- enet_coef_tbl |>
  filter(term != "(Intercept)") |>
  mutate(abs_est = abs(estimate)) |>
  arrange(desc(abs_est)) |>
  slice(1:20)

enet_top
```

Comparing Regression:

```{r}
# ---- 8) Side-by-side performance tables (MAE and RMSE)   now with Elastic Net ----
# Assumes you already computed:
#   pred_ols_train,  pred_ols_test
#   pred_lasso_train, pred_lasso_test
#   pred_ridge_train, pred_ridge_test
#   pred_enet_train,  pred_enet_test
# And (optionally) best_alpha, best_lambda_min from your Elastic Net CV

# Helper: safely carry alpha/lambda for ENet if they exist
alpha_enet  <- if (exists("best_alpha")) best_alpha else NA_real_
lambda_enet <- if (exists("best_lambda_min")) best_lambda_min else NA_real_

# Build TRAIN table
metrics_train <- tibble(
  Model = c("OLS", "LASSO (λ.min)", "Ridge (λ.min)", "Elastic Net")
) |>
  bind_cols(
    dplyr::bind_rows(
      eval_perf(y_train, pred_ols_train),
      eval_perf(y_train, pred_lasso_train),
      eval_perf(y_train, pred_ridge_train),
      eval_perf(y_train, pred_enet_train)
    )
  ) |>
  mutate(Alpha = c(NA, 1.00, 0.00, alpha_enet),
         Lambda = c(NA, NA, NA, lambda_enet))

# Build TEST table
metrics_test <- tibble(
  Model = c("OLS", "LASSO (λ.min)", "Ridge (λ.min)", "Elastic Net")
) |>
  bind_cols(
    dplyr::bind_rows(
      eval_perf(y_test, pred_ols_test),
      eval_perf(y_test, pred_lasso_test),
      eval_perf(y_test, pred_ridge_test),
      eval_perf(y_test, pred_enet_test)
    )
  ) |>
  mutate(Alpha = c(NA, 1.00, 0.00, alpha_enet),
         Lambda = c(NA, NA, NA, lambda_enet))

# Print results
metrics_train

metrics_test
```

Result: On average patients with a high adherance levels had a decrease in creatinine level of -0.09340354 after 48 hours

Main Treatment effect:

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(broom)
  library(stringr)
  library(tibble)
  
})
#checking the variables in OLS
formula(mod_ols)

# OLS: adjusted main effect of treatment (chemo vs reference)

coef_ols <- tidy(mod_ols, conf.int = TRUE) %>%
filter(grepl("^adherence_level", term))
coef_ols
```

```{r}
# Regularized models: report the treatment coefficient (main effect)
grab_coef <- function(m, name) {
b <- as.matrix(coef(m))
if (!name %in% rownames(b)) return(NA_real_)
as.numeric(b[name, 1])
}

main_effects_tbl <- tibble(
Model = c("OLS","LASSO","Ridge","Elastic Net"),
Adherence_To_Protocol = c(
coef(mod_ols)[["adherence_levelHigh adherence"]],
grab_coef(mod_lasso, "adherence_levelHigh adherence"),
grab_coef(mod_ridge, "adherence_levelHigh adherence"),
grab_coef(mod_enet,  "adherence_levelHigh adherence")
)
)
main_effects_tbl
#adherence_levelHigh adherence is the coefficient of the variable adherence_level.
```

Main Effect Associations:

We are comparing the different models using our clinical predictors. The columns in the dataframe are renamed and the glmnet object is converted into a matrix of coefficients. We then filtered for only the variables we want and presented the model results together to display which predictors are significant and stable across models.

```{r}
# --- Clinical (no interactions): OLS + LASSO + RIDGE + ENET ------------------

suppressPackageStartupMessages({
  library(dplyr)
  library(broom)
  library(stringr)
  library(tibble)
})

# 1) OLS clinical table (with CIs and p-values)

ols_tidy <- tidy(mod_ols, conf.int = TRUE)

ols_clinical <- ols_tidy %>%
  filter(
    term == "age" |
    term == "sofa" |
    term == "lactate" |
    term == "map_min" |
    term == "baseline_creat" |
    term == "bun" |
    term == "crp" |
    term == "albumin"
  ) %>%
  arrange(p.value) %>%
  rename(
    beta_ols    = estimate,
    se_ols      = std.error,
    t_ols       = statistic,
    p_ols       = p.value,
    ci_low_ols  = conf.low,
    ci_high_ols = conf.high
  )

# 2) Helper to extract glmnet coefficients ------------------------------------

get_glmnet_coefs <- function(mod_glmnet, keep_regex = NULL, colname = "beta"){
  B <- as.matrix(coef(mod_glmnet))
  
  tab <- tibble(
    term = rownames(B),
    !!colname := as.numeric(B[, 1])
  ) %>%
    filter(term != "(Intercept)")
  
  if (!is.null(keep_regex)) {
    tab <- tab %>% filter(str_detect(term, keep_regex))
  }
  
  tab
}

# Regex to keep only continuous kidney predictors
keep_regex_kidney <- paste(
  c("^age$",
    "^sofa$",
    "^lactate$",
    "^map_min$",
    "^baseline_creat$",
    "^bun$",
    "^crp$",
    "^albumin$"),
  collapse = "|"
)

lasso_tab <- get_glmnet_coefs(mod_lasso,
                              keep_regex = keep_regex_kidney,
                              colname = "beta_lasso")

ridge_tab <- get_glmnet_coefs(mod_ridge,
                              keep_regex = keep_regex_kidney,
                              colname = "beta_ridge")

enet_tab  <- get_glmnet_coefs(mod_enet,
                              keep_regex = keep_regex_kidney,
                              colname = "beta_enet")

# 3) Join everything ----------------------------------------------------------

clin_compare <- ols_clinical %>%
  select(term, beta_ols, se_ols, t_ols, p_ols,
         ci_low_ols, ci_high_ols) %>%
  left_join(lasso_tab, by = "term") %>%
  left_join(ridge_tab, by = "term") %>%
  left_join(enet_tab,  by = "term") %>%
  mutate(
    beta_lasso = coalesce(beta_lasso, 0),
    beta_ridge = coalesce(beta_ridge, 0),
    beta_enet  = coalesce(beta_enet,  0)
  )

data.frame(clin_compare)
```

CHANGEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEE
In order to visualise the main effect association we first had to create an importance metric column from our object clin_compare. We then plotted each predictor with it's strength, with larger bars representing stronger predictors.

```{r}
clin_long <- clin_compare %>%
  select(term, beta_ols, t_ols, beta_lasso, beta_ridge, beta_enet) %>%
  tidyr::pivot_longer(
    cols = -c(term, t_ols),
    names_to = "Model",
    values_to = "value"
  ) %>%
  mutate(
    importance = case_when(
      Model == "beta_ols" ~ abs(t_ols),
      TRUE ~ abs(value)
    ),
    Model = case_when(
      Model == "beta_ols"   ~ "OLS",
      Model == "beta_lasso" ~ "LASSO",
      Model == "beta_ridge" ~ "Ridge",
      Model == "beta_enet"  ~ "Elastic Net"
    )
  )

ggplot(clin_long,
       aes(x = reorder(term, importance), y = importance)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ Model, scales = "free_y") +
  labs(
    x = "Predictor",
    y = "Importance (|t| for OLS; |β| for penalized models)",
    title = "Clinical Predictor Importance by Model"
  ) +
  theme_minimal(base_size = 12)
```
Results:
The OLS model showed predictors such as Lactate, SOFA, Map_min and baseline_creat have a strong association with our response variable. LASSO, Elastic Net and Ridge showed no clear predictors with strong correlation. This could mean that our sample size is too small. 

Tree Methods

Setting up:
```{r}
# Core data/plot
req_pkgs <- c(
  "dplyr", "ggplot2", "tidyr", "readr", "tibble", "gridExtra",
  # Trees
  "rpart", "rpart.plot", "partykit",
  # Random Forests (pick one or use both)
  "ranger",        # fast RF (recommended)
  "randomForest",  # classic RF implementation
  # Gradient Boosting
  "xgboost", "Matrix",  # Matrix for sparse design matrices
  # Model interpretation (optional but handy)
  "vip",   # variable importance plots
  "pdp",   # partial dependence
  "iml"    # ICE/SHAP-like tools (optional)
)

# Install any missing
to_install <- setdiff(req_pkgs, rownames(installed.packages()))
if (length(to_install) > 0) {
  install.packages(to_install, dependencies = TRUE)
}

# Load all (silently)
invisible(lapply(req_pkgs, require, character.only = TRUE))
```

Dividing the dataset into training and testing:

```{r}
# 2) Stratified 70/30 split by treatment
set.seed(42)
split_strat <- function(df, strat_col, p_train = 0.7) {
  idx_tr <- unlist(tapply(seq_len(nrow(df)), df[[strat_col]], function(ix) {
    sample(ix, size = floor(p_train * length(ix)))
  }))
  list(train = sort(idx_tr), test = setdiff(seq_len(nrow(df)), idx_tr))
}

sp <- split_strat(kidney_full, strat_col = "adherence_level", p_train = 0.7)
train <- kidney_full[sp$train, , drop = FALSE]
test  <- kidney_full[sp$test,  , drop = FALSE]
```

```{r}
# 3) Drop columns that should NOT enter models
#    - patient_id: identifier only
#    - high_response: binary version of the outcome (leakage if modeling response_percent)
#    - baseline_tumor_mm, post_tumor_mm: strongly deterministically related to response_percent
drop_cols <- intersect(
  names(train),
  c("patient_id", "site", "protocol_adherence", "baseline_creat", "creatinine_change")
)
train_nopii <- dplyr::select(train, -dplyr::all_of(drop_cols))
test_nopii  <- dplyr::select(test,  -dplyr::all_of(drop_cols))

# 4) Check outcome presence
#stopifnot("response_percent" %in% names(train_nopii))
```

```{r}
# 5) Build a consistent design (for glmnet / xgboost, etc.)
#    Use TRAIN to “freeze” factor levels and dummy columns
f_ols   <- delta_creatinine_48h ~ .
ols_tmp <- lm(f_ols, data = train_nopii)
ols_terms <- terms(ols_tmp)

# Model matrices (no intercept column)
X_train <- model.matrix(ols_terms, data = train_nopii)[, -1, drop = FALSE]
X_test  <- model.matrix(ols_terms, data = test_nopii)[,  -1, drop = FALSE]
y_train <- train_nopii$delta_creatinine_48h
y_test  <- test_nopii$delta_creatinine_48h
```

```{r}
# 6) Small helpers for later evaluations
mae  <- function(y, yhat) mean(abs(y - yhat))
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))
eval_perf <- function(y_true, y_pred) {
  tibble::tibble(MAE = mae(y_true, y_pred), RMSE = rmse(y_true, y_pred))
}
```


Interaction Plots: 

For the plots below we are using delta_creatinine_48h as our outcome, this is plotted using the average (mean) creatinine change per group. Adherence level as our grouping variable. Lactate, SOFA and Map_min are used as the continuous predictors for the three plots. Each predictor is aggregated into 4 breaks for visualisation purposes. 

```{r}
agg_inter <- kidney_full %>%
  mutate(age = cut(age, breaks = 4)) %>%
  group_by(age, adherence_level) %>%
  summarise(
    mean_creat_change = mean(delta_creatinine_48h, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(
  agg_inter,
  aes(x = age,
      y = mean_creat_change,
      color = adherence_level,
      group = adherence_level)
) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3) +
  labs(
    title = "Interaction Plot (Kidney Dataset)",
    subtitle = "Parallel lines → effect of age across adherence levels",
    x = "Age",
    y = "Mean Creatinine Change",
    color = "Adherence Level"
  ) +
  theme_minimal(base_size = 14)
```
This plot indicates that there is a much stronger adherence dependent variation in mean creatinine change of patients in relation to age. There was a much higher change in patients in the low adherance group when compared to their age group counterparts who had a high adherance. 

```{r}
agg_inter <- kidney_full %>%
  mutate(lactate_group = cut(lactate, breaks = 4)) %>%
  group_by(lactate_group, adherence_level) %>%
  summarise(
    mean_creat_change = mean(delta_creatinine_48h, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(
  agg_inter,
  aes(x = lactate_group,
      y = mean_creat_change,
      color = adherence_level,
      group = adherence_level)
) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3) +
  labs(
    title = "Interaction Plot (Kidney Dataset)",
    subtitle = "Parallel lines → effect of lactate similar across adherence levels",
    x = "Lactate (Grouped)",
    y = "Mean Creatinine Change",
    color = "Adherence Level"
  ) +
  theme_minimal(base_size = 14)

```

There appears to be a strong correlation between lactate and mean creatinine change. There also seems to be an adhearance dependent variability between the patients. High adhearance seemed to plateau between 0.5 and 0.75 mg/L change where as the low adhearance continued to change much further. This could indicate that the low adhearance group struggled to regulate the bodies inflammatory response or were more susceptible to develop a complication in this cohort. 

```{r}
agg_inter2 <- kidney_full %>%
  mutate(sofa_group = cut(sofa, breaks = 4)) %>%
  group_by(sofa_group, adherence_level) %>%
  summarise(
    mean_creat_change = mean(delta_creatinine_48h, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(
  agg_inter2,
  aes(x = sofa_group,
      y = mean_creat_change,
      color = adherence_level,
      group = adherence_level)
) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3) +
  labs(
    title = "Interaction Plot (Kidney Dataset)",
    subtitle = "Parallel lines → effect of SOFA similar across adherence levels",
    x = "SOFA (Grouped)",
    y = "Mean Creatinine Change",
    color = "Adherence Level"
  ) +
  theme_minimal(base_size = 14)


agg_inter3 <- kidney_full %>%
  mutate(mapmin_group = cut(map_min, breaks = 4)) %>%
  group_by(mapmin_group, adherence_level) %>%
  summarise(
    mean_creat_change = mean(delta_creatinine_48h, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(
  agg_inter3,
  aes(x = mapmin_group,
      y = mean_creat_change,
      color = adherence_level,
      group = adherence_level)
) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3) +
  labs(
    title = "Interaction Plot (Kidney Dataset)",
    subtitle = "Parallel lines → effect of MAP_MIN similar across adherence levels",
    x = "Mean Arterial Pressure (Grouped)",
    y = "Mean Creatinine Change",
    color = "Adherence Level"
  ) +
  theme_minimal(base_size = 14)
```

Running a tree and Pruning via 1-SE rule:
```{r}
# Grow a CART regression tree
kidney_tree <- rpart(
  delta_creatinine_48h ~ .,
  data = train_nopii,
  method = "anova",
  control = rpart.control(cp = 0.001, maxdepth = 8, minsplit = 30)
)

# Prune via 1-SE rule
printcp(kidney_tree)
```

```{r}
best_row <- which.min(kidney_tree$cptable[,"xerror"])
xerr_min <- kidney_tree$cptable[best_row, "xerror"]
xstd_min <- kidney_tree$cptable[best_row, "xstd"]
cp_1se   <- kidney_tree$cptable[kidney_tree$cptable[,"xerror"] <= xerr_min + xstd_min, "CP"][1]
kidney_pruned <- prune(kidney_tree, cp = cp_1se)

# Human-readable rules (if/else)
rpart.plot::rpart.rules(kidney_pruned, style = "tallw")
```
Visualising grown tree. Full tree is too complex and branched to visualise clearly, therefore the pruned tree is plotted instead.
```{r}
# Visualize grown and pruned trees
rpart.plot(kidney_pruned, type = 2, extra = 101, box.palette = "Blues",
           main = "Kidney dataset tree (Pruned 1-SE)")
```

```{r}
pred_tr <- predict(kidney_pruned, newdata = train_nopii)
pred_te <- predict(kidney_pruned, newdata = test_nopii)

perf <- dplyr::bind_rows(
  tibble(Split="Train", MAE=mae(train_nopii$delta_creatinine_48h, pred_tr),
         RMSE=rmse(train_nopii$delta_creatinine_48h, pred_tr)),
  tibble(Split="Test",  MAE=mae(test_nopii$delta_creatinine_48h,  pred_te),
         RMSE=rmse(test_nopii$delta_creatinine_48h,  pred_te))
)
perf
```

```{r}
ggplot(
  tibble(truth = test_nopii$delta_creatinine_48h, pred = pred_te),
  aes(truth, pred)
) +
  geom_point(alpha = 0.35) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  labs(x = "Observed Creatinine change", y = "Tree prediction",
       title = "Pruned tree: observed vs predicted (test)") +
  theme_minimal()
```

Bootstrap Aggregating

Setting up test and train for dataset:
```{r}
set.seed(2025)

library(dplyr)

# 1) Train/test split (70/30) -------------------------
n <- nrow(kidney_full)
idx_train <- sample(seq_len(n), size = 0.7 * n)

train <- kidney_full[idx_train, ]
test  <- kidney_full[-idx_train, ]

# 2) True outcomes ------------------------------------
y_train <- train$delta_creatinine_48h
y_test  <- test$delta_creatinine_48h
```


Random Forests:

Code is adjusted to fit the package randomForest.
```{r}
library(randomForest)

rf_fit <- randomForest(
  delta_creatinine_48h ~ . 
    - patient_id 
    - site 
    - creatinine_change
    - protocol_adherence
    - baseline_creat,
  data       = train,
  ntrees  = 500,
  mtry       = floor(sqrt(ncol(train) - 6)),  # approx: drop 5 predictors + outcome
  importance = TRUE
)
```


```{r}
# Predictions ------------------------------------------------------

rf_pred_train <- predict(rf_fit, newdata = train)
rf_pred_test  <- predict(rf_fit, newdata = test)

# Performance ------------------------------------------------------

rf_perf_train <- eval_perf(y_train, rf_pred_train)
rf_perf_test  <- eval_perf(y_test,  rf_pred_test)

rf_perf_train
```

```{r}
rf_perf_test
```

XGboost:

```{r}
library(xgboost)

# 1) Build predictor matrices -------------------------------------

cols_to_drop <- c(
  "patient_id",
  "site",
  "creatinine_change",
  "protocol_adherence",
  "baseline_creat",
  "delta_creatinine_48h")

X_train <- model.matrix(
  ~ .,
  data = train |> select(-all_of(cols_to_drop))
)

X_test <- model.matrix(
  ~ .,
  data = test  |> select(-all_of(cols_to_drop))
)

# Optional: check dimensions
dim(X_train)
```
```{r}
dim(X_test)
```

```{r}
# 2) Convert to DMatrix -------------------------------------------

dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest  <- xgb.DMatrix(data = X_test,  label = y_test)

# 3) Basic XGBoost hyperparameters --------------------------------

xgb_params <- list(
  objective        = "reg:squarederror",
  eta              = 0.05,   # learning rate
  max_depth        = 4,
  subsample        = 0.7,    # row subsampling
  colsample_bytree = 0.7     # column subsampling
)

# 4) Train XGBoost model ------------------------------------------

xgb_fit <- xgb.train(
  params    = xgb_params,
  data      = dtrain,
  nrounds   = 300,
  watchlist = list(train = dtrain),
  verbose   = 0
)

# 5) Predictions ---------------------------------------------------

xgb_pred_train <- predict(xgb_fit, dtrain)
xgb_pred_test  <- predict(xgb_fit, dtest)

# 6) Performance ---------------------------------------------------

xgb_perf_train <- eval_perf(y_train, xgb_pred_train)
xgb_perf_test  <- eval_perf(y_test,  xgb_pred_test)

xgb_perf_train
```

```{r}
xgb_perf_test
```


```




